---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data
The data will be read into a variable named 'data'.
```{r}
data <- read.csv(unz("activity.zip", 'activity.csv'))
```

Import libraries.
```{r echo=TRUE, message=FALSE}
library(dplyr)
library(xtable)
library(ggplot2)
```

## What is mean total number of steps taken per day?
**1. Total number of steps**
Total steps per day is:
```{r echo=TRUE, results='asis'}
total_daily_steps <- summarise(
  group_by(data, date),
  daily_steps = sum(steps, na.rm = TRUE)
  )
print(xtable(total_daily_steps), type='html')
```


**2. Histogram of number of steps per day**
```{r}
hist(total_daily_steps$daily_steps, main = "Histogram of Total Steps per Day", xlab="steps", ylab="frequency")
```

**3. Mean and median of steps per day**
```{r echo=TRUE}
avg_steps <- mean(total_daily_steps$daily_steps)
median_steps <- median(total_daily_steps$daily_steps)
```
- Average steps per day is `r avg_steps`
- Median steps per day is `r median_steps`

## What is the average daily activity pattern?
**1. Make a time series plot, of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)**
```{r}
steps_by_interval <- summarise(
  group_by(data, interval),
  steps = mean(steps, na.rm=TRUE)
)

qplot(interval, steps, data=steps_by_interval, geom=c('line'))
```

**2. Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?**
```{r}
max_interval <- filter(steps_by_interval, steps == max(steps_by_interval$steps))$interval[1]
```
The interval with the highest average steps is `r max_interval`.

## Imputing missing values
```{r}
missing_values <- dim(filter(data, is.na(steps)))[1]
```

**1. Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NAs)**  
- There are `r missing_values` missing values in the dataset.

**2. Devise a strategy for filling in all of the missing values in the dataset.**  
A strategy to fill in all of the missing values is to use the daily average of that interval.

**3. Create a new dataset that is equal to the original dataset but with the missing data filled in.**
```{r}
data_fillna <- merge(data, steps_by_interval, by = c('interval'))
data_fillna$steps <- data_fillna$steps.x
data_fillna$steps[is.na(data_fillna$steps.x)] <- data_fillna$steps.y[is.na(data_fillna$steps.x)]

data_fillna <- data_fillna[, c('steps', 'date', 'interval')]
```

**4. Make a histogram of the total number of steps taken each day and Calculate and report the mean and median total number of steps taken per day**

```{r}
total_daily_fillna <- summarise(group_by(data_fillna, date),
                                steps = sum(steps)
                                )
avg_daily_fillna <- mean(total_daily_fillna$steps)
median_daily_fillna <- median(total_daily_fillna$steps)

hist(total_daily_fillna$steps, main='Histogram of Daily Steps (fillna)', xlab = 'steps', ylab = 'frequency')
```

- The average daily steps after fill na becomes: `r avg_daily_fillna`
- The median daily steps after fill na becomes: `r median_daily_fillna`
- Imputing the missing values had more impact on the mean than the median, and lowered the variance in estimating daily average steps.

## Are there differences in activity patterns between weekdays and weekends?
**1. Creating a new weekend flag variable **  
```{r}
data_fillna$weekday <- 'weekday'
data_fillna$weekday[weekdays(as.Date(data_fillna$date)) == 'Saturday'] <- 'weekend'
data_fillna$weekday[weekdays(as.Date(data_fillna$date)) == 'Sunday'] <- 'weekend'
```

**2. Panel plot of the intervals**  
```{r}
data_by_weekday <- summarise(
  group_by(data_fillna, weekday, interval),
  steps = mean(steps)
  )
qplot(interval, steps, facets = .~weekday, data=data_by_weekday, geom=c('line'))
```
